\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amssymb}

\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
\newcommand{\code}[1]{\texttt{#1}}

\title{Artificial Economies for Learning Agents:\\
Testing Whether Institutions Stay Robust as Agent Capability Changes}
\author{
  \todo{Author name}\\
  \todo{Institution / Department}\\
  \todo{Email}
}
\date{Draft v0.2 -- \today}

\begin{document}
\maketitle

\begin{abstract}
This paper introduces a code-first experimental platform for studying whether
economic mechanisms remain robust when the strategic agents inside them are
adaptive learners rather than equilibrium solvers. The platform implements a
shared \code{World}/\code{Agent}/\code{Institution} interface, repeated
multiseed evaluation, exploitability tests, and interchangeable mind classes
ranging from random and tabular Q-learning agents to PyTorch DQN, PPO,
independent-DQN, and centralized-critic agents. The first validated environment
is a repeated two-firm pricing arena with six regulatory mechanisms. In this
world, price caps robustly reduce exploitability across mind classes, but their
effect on collusion is metric- and capability-dependent: stronger DQN-family
learners can sit effectively at the cap while preserving high profit through
the quantity margin. The second environment, Resource Island, extends the same
interface to a spatial inventory economy and reveals a different lesson:
institution conclusions are only meaningful when the world actually exercises
the institutional channel. Early Resource Island versions mechanically
under-tested property rights and trade price controls; a hardened v1 now
creates contested resources, unequal trades, and specialization pressure. Three
additional environments extend the ladder: Auction House provides a benchmarked
auction-design testbed, Public Goods exposes common-pool extraction and
contribution incentives, and Labor Market tests learned reporting under
deferred acceptance. The current contribution is both empirical and
methodological: it reports initial evidence that institutional robustness can
change with learner capability, and it documents the validation discipline
needed to distinguish real economic findings from inactive-mechanism artifacts.
\end{abstract}

\section*{First-Page Build Map}

This draft is intentionally structured as the arXiv-facing version of the
project, with the parts that already have evidence separated from the parts
that still need final runs or prose.

\begin{center}
\begin{tabular}{p{0.2\linewidth}p{0.36\linewidth}p{0.34\linewidth}}
\toprule
Part & What It Must Do & Current Status \\
\midrule
Motivation & State the core question: do mechanisms stay robust under a
capability ladder of learning agents? & Ready to draft. \\
Platform & Define \code{World}, \code{Agent}, \code{Institution}, metrics,
manifests, parity tests, and output validation. & Built and tested. \\
Pricing Arena & Main mature result: collusion, welfare, exploitability, and
capability dependence in a repeated pricing game. & Full n=20 outputs exist;
fixed independent-DQN is merged into the main table. \\
Resource Island & Non-pricing stress test: spatial resources, inventory,
trade, property, redistribution, reputation. & v0 diagnosed; v1 and cross-mind
full outputs validated. \\
Auction House & Third world: sealed-bid auctions with theory benchmarks. &
Mechanics, benchmarks, Q-learning full run, P.6 smoke, and full P.6 queue are
active. \\
Public Goods & Common-pool resource world with contribution, extraction,
sanctions, matching, information restriction, reputation, and taxes. & Q-learning
full run and P.6 smoke exist; full P.6 queue is active. \\
Labor Market & Two-sided matching world with learning workers and fixed
employer preferences under deferred acceptance. & Q-learning full run and P.6
smoke exist; full P.6 queue is active. \\
Synthesis & Convert runs into thesis claims: which institutional effects
generalize, which are world-specific, and which are capability-sensitive. &
Preliminary five-world smoke synthesis exists; final claims wait on current
full P.6 queues. \\
Limitations & Be explicit about inactive institutions, finite compute, simple
worlds, and non-equilibrium learned behavior. & Ready to draft. \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Stop condition for the first complete paper draft: after the current Auction
House, Public Goods, and Labor Market full P.6 queues validate cleanly, freeze
the build-out temporarily and write the cross-world synthesis before adding
another environment or LLM-agent layer.

\section{Introduction}

Economic mechanisms are usually evaluated against a model of strategic
behavior: equilibrium prices, truthful bidding, efficient allocation, or some
other theoretical benchmark. That approach is powerful, but it leaves open a
practical question that becomes more important as deployed systems are
increasingly populated by adaptive software agents: what happens when the
agents do not solve the theorist's equilibrium, but instead learn from repeated
interaction?

This paper studies that question with artificial economies: small, controlled
economic worlds in which institutions can be swapped while the learning
architecture of the agents is varied. The goal is not to simulate a whole
economy. The goal is to create environments that are simple enough to audit but
rich enough to expose failures of institutional robustness. A mechanism is
treated as robust only if it performs well across seeds, across learner classes,
and under adversarial or exploitability-style checks.

The central research question is:

\begin{quote}
When agent capability increases from random behavior to tabular Q-learning and
deep reinforcement learning, do the same economic institutions still reduce
collusion, exploitability, inequality, and inefficient strategic behavior?
\end{quote}

The current evidence already gives a nontrivial answer in the pricing world.
Price caps reduce exploitability for every tested mind class, but their
collusion effect depends on the operational metric and on learner capability.
In the full Pricing Arena comparison, tabular Q-learning under no regulation has
average price 5.49, total profit 365.34, and exploitability 95.69. Under a price
cap, average price falls to 4.73 and exploitability falls to 40.32. For
DQN-family learners, however, the cap can create a different pattern: the agents
sit near the ceiling, with low price dispersion, while preserving high
profit-normalized collusion through quantity and demand effects. That is the
kind of result this platform is meant to surface: a regulation can look strong
under one behavioral lens and weaker under another.

\section{Related Work}

This project sits at the intersection of algorithmic collusion, reinforcement
learning, and mechanism design. The Pricing Arena is motivated by work showing
that learning algorithms can sustain supra-competitive outcomes in repeated
pricing environments \citep{calvano2020,calvano2021}. The exploitability
protocol is related to the broader idea that a learned policy should not only be
evaluated by its on-policy average outcome, but also by its brittleness against
strategic deviation.

The mind ladder follows standard reinforcement-learning baselines:
tabular Q-learning and deep Q-networks \citep{sutton2018,mnih2015}, PPO
\citep{schulman2017}, and independent multi-agent Q-learning in the spirit of
\citet{tan1993}. The Auction House environment connects the platform back to
classic mechanism-design results: truthful bidding in second-price auctions
\citep{vickrey1961}, revenue and allocation tradeoffs in auction design
\citep{myerson1981,milgrom2004}, and strategic bid shading in first-price
auctions.

The Public Goods and Resource Island environments connect the platform to
common-pool resource and social-dilemma settings, where individually profitable
extraction can destroy shared surplus and where sanctions, reputation, and
matching contributions are standard institutional responses. The Labor Market
environment connects to two-sided matching and deferred acceptance, where
truthful revelation and stability provide unusually sharp benchmark concepts.

\todo{Tighten this section with exact citations for Calvano et al. 2021,
Asker--Fershtman--Pakes, and Banchio--Skrzypacz once the literature pass is
done.}

\section{Platform}

\subsection{Interfaces}

The codebase is organized around three abstractions:

\begin{itemize}
  \item A \code{World} defines state transitions, observations, metrics, and
  rendering state.
  \item An \code{Agent} or mind maps observations to actions and updates from
  rewards and next observations.
  \item An \code{Institution} modifies rules, rewards, constraints, or
  information inside a world.
\end{itemize}

This separation is not cosmetic. It makes it possible to ask whether a result is
about a world, an institution, or a learner. It also creates a regression test:
if a new world requires rewriting the shared learning interface, then the
platform abstraction has failed.

\subsection{Validation Discipline}

The platform treats a result as reportable only after three checks. First, the
mechanics must be unit-tested: allocation, movement, payment, trade, collision,
or reward rules must match hand-computed cases. Second, the experiment must be
replicated over seeds, with per-seed and aggregate CSV outputs. Third, the
institution must actually activate. If a property-rights rule never faces a
non-owner gather attempt, or a trade price control can never bind because every
trade is one-for-one, then the result is not evidence about that institution. It
is evidence about the world design.

This distinction matters for Resource Island. The initial version ran and
passed tests, but three institutions were mechanically under-exercised. Rather
than report those as null effects, the project added diagnostic counters and a
v1 pressure configuration with contested resources and unequal trades.

\subsection{Mind Ladder}

The current mind ladder includes random agents, tabular Q-learning, PyTorch DQN,
PyTorch PPO, independent-DQN, and a centralized-critic scaffold. The DQN
implementation uses replay, a target network, Huber TD loss, and a
DeepMind-style centered RMSProp optimizer. PPO uses categorical actions, GAE,
clipped policy loss, value loss, entropy regularization, and minibatch epochs.

An important implementation issue was discovered and fixed during the project:
the original independent-DQN condition behaved like a DQN alias in several
tables. The corrected implementation now uses separate per-agent learners,
decorrelated child random streams, separate replay and exploration randomness,
and scoped PyTorch initialization. Fixed independent-DQN replacement rows are
now merged for Pricing Arena and Resource Island, and the same corrected
condition is used in the active Auction House, Public Goods, and Labor Market
P.6 queues.

\section{Environment 1: Pricing Arena}

\subsection{World}

Pricing Arena is a repeated two-firm pricing market. Each firm chooses a price
from a discrete grid. Consumers choose between firms according to a noisy logit
demand model. Firms earn profit from demand net of costs, and learning agents
update from repeated play. The implemented mechanisms are no regulation, price
cap, high-price tax, random audits, anti-collusion penalty, and demand shocks.

\subsection{Metrics}

The main metrics are average price, total profit, consumer surplus, welfare,
exploitability, and two collusion indexes. The historical project metric is a
price-normalized proxy. The literature-comparable metric is a profit-normalized
index:

\begin{equation}
  \mathrm{CI}_{profit}
  =
  \frac{\pi_{observed} - \pi_{Nash}}
       {\pi_{monopoly} - \pi_{Nash}}.
\end{equation}

Both are reported because the divergence between them is empirically
informative.

\subsection{Main Result}

Table~\ref{tab:pricing-cap} shows a compact version of the mature Pricing Arena
result for Q-learning and random baselines. Price caps reduce exploitability and
raise welfare relative to the no-regulation case in these rows.

\begin{table}[h]
\centering
\begin{tabular}{llrrrrr}
\toprule
Mind & Mechanism & Price & Profit & Welfare & Profit CI & Exploitability \\
\midrule
Q-learning & none & 5.49 & 365.34 & 810.43 & 0.475 & 95.69 \\
Q-learning & price cap & 4.73 & 342.42 & 829.99 & 0.424 & 40.32 \\
Random & none & 5.51 & 277.92 & 785.05 & 0.282 & 54.62 \\
Random & price cap & 4.32 & 261.89 & 819.83 & 0.247 & 22.84 \\
\bottomrule
\end{tabular}
\caption{Selected Pricing Arena full-run results from the Phase 3 comparison
output.}
\label{tab:pricing-cap}
\end{table}

The more interesting finding appears when learner capability increases. In the
Phase 3 table, price caps reduce exploitability across tested minds, but
collusion is metric-sensitive for stronger learners. DQN-family agents can sit
effectively at the cap, with low price dispersion, while maintaining high total
profit. The current diagnosis is that the cap constrains the price margin but
does not eliminate the quantity/profit channel.

\todo{Add a full table comparing DQN, PPO, centralized-critic, and corrected
independent-DQN under price cap versus no regulation.}

\section{Environment 2: Resource Island}

\subsection{World}

Resource Island is a bounded-grid gather/trade economy. Agents move, gather food
and wood, maintain energy, accumulate inventories, and can die from starvation.
The world contains property rights, redistribution, trade price controls, and
reputation institutions. It was built to test whether the platform abstractions
work outside a two-firm pricing game.

\subsection{What v0 Taught}

The first full Resource Island runs were mechanically valid but economically
under-activated. In the initial full run, successful trade was sparse and some
institutions did not bind. Follow-up replay showed that property rights almost
never faced behavioral pressure: even when claims existed, non-owner
opportunities to gather from claimed resource cells were extremely rare.
Similarly, trade price controls could not bind because the original trade
protocol only allowed one-for-one exchange.

This is an important negative result for the platform: a world can be runnable,
tested, and statistically replicated while still failing to exercise the
institution it is supposed to study.

\subsection{v1 Hardening}

Resource Island v1 adds four changes:

\begin{enumerate}
  \item contested resource layouts, so multiple agents have reason to access the
  same cells;
  \item unequal trade offers, so price controls can actually bind;
  \item specialization pressure through heterogeneous resource preferences and
  starting inventories;
  \item activation diagnostics for trade attempts, blocked trades, property
  opportunities, property violations, and institution blocks.
\end{enumerate}

The v1 validation run verified that the economic channels were active, and the
subsequent n=20 full run preserves that activation. Table~\ref{tab:resource-v1}
shows the medium validation values used as the pre-full-run gate.

\begin{table}[h]
\centering
\begin{tabular}{lrrrrr}
\toprule
Institution & Welfare & Survival & Trades & Trade Blocks & Property Opps. \\
\midrule
none & 1.095 & 0.933 & 6.034 & 0.000 & 0.000 \\
property rights & 1.078 & 0.923 & 5.790 & 0.000 & 17.734 \\
trade price controls & 0.875 & 0.898 & 0.000 & 5.779 & 0.000 \\
reputation & 1.313 & 0.927 & 6.867 & 0.000 & 0.000 \\
\bottomrule
\end{tabular}
\caption{Resource Island v1 validation results.}
\label{tab:resource-v1}
\end{table}

The interpretation is straightforward. Price controls now bind because the
world permits unequal trades. Reputation increases successful trade and welfare
in the validation setting. Property-right opportunities are present, although
the welfare effect is small in this medium run. The full n=20 v1 run confirms
the activation pattern: successful trade appears under no institution, price
controls bind and suppress trade, property-right opportunities are measured,
and reputation produces the strongest welfare gains in this pressure setup.

\subsection{Cross-Mind Resource Island}

The Resource Island Phase 3 full table gives a different pattern from Pricing
Arena. Neural/MARL minds find higher survival and welfare policies, but they do
not learn the intended trade economy under the current reward and training
horizon. On the no-institution condition, Q-learning has survival 0.969,
welfare 1.005, and trade count 0.294. DQN has survival 0.987 and welfare 1.257
but trade count 0.000. PPO and centralized-critic show the same qualitative
pattern.

This should not be over-interpreted as proof that stronger agents strategically
avoid trade. The trade-attempt diagnostics suggest exploration and coordination
failure: DQN and centralized-critic make rare attempts, PPO makes essentially no
attempts, and attempts tend to be inventory-blocked. The safe claim is that
these stronger learners find high-survival policies without discovering
successful trade under the current Resource Island setup.

\todo{Add a compact Resource Island cross-mind table with Q-learning, DQN, PPO,
fixed independent-DQN, and centralized-critic on the no-institution row.}

\section{Environment 3: Auction House}

\subsection{World}

Auction House is a single-item sealed-bid auction world with independent private
values, discrete bids, deterministic tie-breaking, first-price and second-price
payment rules, and optional reserve prices. Each bidder observes its own
valuation bin and chooses a bid. The world reports seller revenue, bidder
surplus, total welfare, allocative efficiency, regret, overbidding, underbidding,
and distance from benchmark bid functions.

\subsection{Benchmarks}

The second-price benchmark is truthful bidding. The first-price benchmark is
bid shading under symmetric independent private values, with discrete-grid
regret checks used where the continuous benchmark is too coarse. Reserve-price
variants evaluate the revenue-efficiency tradeoff against the no-reserve
benchmark.

\subsection{Validation}

The Auction House validation run is economically coherent but not yet final
evidence. Second-price learning moves toward the truthful benchmark, first-price
learning shows underbidding and bid shading, and reserve prices increase revenue
while changing allocative efficiency. Table~\ref{tab:auction-validation}
summarizes the current validation run.

\begin{table}[h]
\centering
\begin{tabular}{lrrrr}
\toprule
Scenario & Revenue & Welfare & Efficiency & Underbid Rate \\
\midrule
second price & 3.373 & 6.216 & 0.747 & 0.381 \\
first price & 3.481 & 6.331 & 0.768 & 0.774 \\
second price + reserve & 4.031 & 6.098 & 0.814 & 0.366 \\
\bottomrule
\end{tabular}
\caption{Auction House validation results.}
\label{tab:auction-validation}
\end{table}

\todo{Replace or extend this table with the active full P.6 Auction House
ladder runs once DQN, PPO, fixed independent-DQN, and centralized-critic
validate. Add learned bid curves against truthful and first-price shaded
reference curves.}

\section{Environment 4: Public Goods}

\subsection{World}

Public Goods is a repeated common-pool resource environment. Agents choose
between contributing to the pool, extracting from it, or remaining inactive.
Extraction pays private reward but depletes the shared stock; contribution costs
the contributor but helps sustain the pool. When requested extraction exceeds
available stock, extraction is rationed proportionally. Deterministic
regeneration makes the sustainability channel easy to audit.

The implemented institutions are no intervention, a penalty schedule,
contribution matching, reputation rewards, information restriction, and a
generic tax-and-redistribution schedule. The main metrics are welfare,
sustainability, total contribution, extraction relative to regeneration,
inequality, collapse diagnostics, and tax revenue where applicable.

\subsection{Q-Learning Full Run}

The n=20 tabular Q-learning full run confirms the intended commons pressure. In
the baseline, agents mostly extract and contribute little: sustainability is
0.0892 and total contribution is 0.1418. Contribution matching improves both
welfare and sustainability in this setting, reaching welfare 2.0808 and
sustainability 0.1048. Reputation is a strong reward-shaping intervention, with
welfare 9.5750, while the tax schedule mainly changes redistribution/accounting
at the tested rates.

The effect validator classifies the tax schedule as mostly reward/accounting
under this configuration, while penalty, contribution matching, reputation, and
information restriction change state metrics relative to baseline. This is a
diagnostic classification, not yet a final welfare ranking.

\subsection{Cross-Mind Status}

Public Goods is P.6 smoke-validated and currently running full P.6 ladder jobs.
The question for the full table is whether stronger learners worsen commons
collapse by extracting more efficiently, or improve sustainability by
discovering contribution and institution-mediated cooperation.

\todo{Insert the full Public Goods P.6 comparison after the active remote queue
validates.}

\section{Environment 5: Labor Market}

\subsection{World}

Labor Market is a two-sided matching environment. Workers are learning agents;
employers have fixed preferences. Each worker reports a top choice, the world
constructs reported preferences, and matching is resolved using deferred
acceptance. The world reports match rate, total welfare, truthful-report rate,
stability, blocking-pair diagnostics, and manipulation-gain diagnostics.

This world is intentionally asymmetric. Unlike Pricing Arena, Resource Island,
Public Goods, and Auction House, not every economic actor is a learning agent.
That asymmetry is the main interface test for the capability ladder: DQN, PPO,
independent-DQN, and centralized-critic operate only over the worker side while
employers remain fixed preference holders.

\subsection{Q-Learning Full Run}

The n=20 tabular Q-learning full run is substantially more stable than the
short smoke run. Match rate remains \code{1.0}, while
\code{stability\_mean=0.9508}, \code{truthful\_report\_rate\_mean=0.7859}, and
\code{total\_welfare\_mean=3.6393}. The report-top action space creates
truthfulness variation, but worker-proposing deferred acceptance still produces
mostly stable matchings under the current random-preference setup.

Fixed benchmark cases clarify the interpretation. Under worker-proposing
deferred acceptance, workers should not have profitable report deviations in
the canonical strategy-proof setting. Future manipulation tests therefore need
to target the non-proposing side, change the mechanism, or add information and
commitment frictions rather than treating worker misreports under standard
deferred acceptance as the main expected failure mode.

\subsection{Cross-Mind Status}

Labor Market is P.6 smoke-validated and currently running full P.6 ladder jobs.
The active full runs test whether stronger learners change truthful-report
rates, stability, and welfare in the asymmetric worker-only learning setup.

\todo{Insert the full Labor Market P.6 comparison after the active remote queue
validates.}

\section{Cross-World Synthesis}

The emerging thesis is not that one institution always works. The emerging
thesis is that institutional robustness must be evaluated jointly with learner
capability and with activation diagnostics.

The Pricing Arena result shows capability-sensitive institutional effects:
price caps reduce exploitability broadly, but stronger learners can preserve
profit under the cap through a channel that a price-only collusion metric would
miss. Resource Island shows the complementary failure mode: if the world design
does not put an institution under pressure, then the absence of an effect is not
a mechanism result. Auction House adds theory-anchored bidding benchmarks.
Public Goods adds a sustainability and commons-collapse channel. Labor Market
adds a matching-stability and truthfulness channel.

The preliminary five-world synthesis scaffold already exists, but three of its
worlds currently use smoke-scale neural/MARL rows. Those rows are useful for
checking the pipeline, not for final claims. The active full P.6 queues are
therefore the current paper blocker: once Auction House, Public Goods, and
Labor Market validate, the synthesis can be rewritten around five full
capability-ladder tables instead of two full tables plus three smoke gates.

\todo{After current full P.6 queues land, write this section as claims rather
than a roadmap. Required inputs: final Pricing Arena table, final Resource
Island table, final Auction House P.6 table, final Public Goods P.6 table, final
Labor Market P.6 table, and regenerated cross-world synthesis outputs.}

\section{Limitations}

Several limitations are deliberate. The environments are small and stylized.
They are designed for auditability rather than realism. The deep-RL agents are
structured-observation MLP agents rather than large-scale policies. The
Resource Island trade protocol is still simple, and the current v1 pressure
configuration is an activation test rather than a realistic market. Auction
House currently has benchmarked sealed-bid full runs and smoke-tested clock and
information variants. Public Goods is still a stylized common-pool model rather
than an empirical commons. Labor Market currently uses worker-side learning
under deferred acceptance, so it tests one asymmetric matching channel rather
than matching-market design generally.

The most important limitation is interpretive. A clean CSV does not imply a
clean economic result. Every institution must be checked for activation. Every
capability comparison must be checked for implementation artifacts. The
independent-DQN alias issue is a concrete example: the table was statistically
valid as an output file, but not valid as evidence for a distinct mind until the
implementation was corrected and rerun.

\section{Reproducibility}

The repository records validation status in \code{ROADMAP\_STATUS.md}. Core
outputs are stored under \code{outputs/}. Major runners include:

\begin{itemize}
  \item \code{run\_multiseed.py} for Pricing Arena replicated runs;
  \item \code{run\_exploitability.py} for frozen-policy adversarial evaluation;
  \item \code{run\_phase3\_full.py} and \code{validate\_phase3\_full.py} for
  full Phase 3 Pricing Arena checks;
  \item \code{run\_resource\_island\_smoke.py} for Resource Island smoke,
  validation, and full runs;
  \item \code{run\_auction\_house\_smoke.py} for Auction House runs;
  \item \code{run\_public\_goods\_smoke.py} for Public Goods runs;
  \item \code{run\_labor\_market\_smoke.py} for Labor Market runs;
  \item \code{build\_world\_mind\_comparison.py} and
  \code{build\_cross\_world\_synthesis.py} for ladder comparisons and
  cross-world synthesis.
\end{itemize}

The test suite currently includes unit, integration, and output-validation
coverage for the core abstractions, Pricing Arena, Resource Island, Auction
House, Public Goods, Labor Market, and deep-RL/MARL minds.

\section{Conclusion}

This draft reports a first version of an artificial-economies platform for
testing institutional robustness under learning agents. The mature Pricing
Arena result already shows why the question matters: exploitability and
collusion can move differently under the same institution, and stronger learners
can expose channels that weaker learners do not. Resource Island, Auction
House, Public Goods, and Labor Market extend the project beyond pricing games.
Their role is not merely to add more checkboxes, but to test whether the
patterns found in Pricing Arena survive when the economic structure changes.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
